Different forms of uncertainty differently affect the evolution of social learning

### 2022-08-13 ms.tex
revise, extend, restructure supplement

By disambiguating and organizing various operationalizations of
uncertainty, we developed a more nuanced theoretical understanding of which forms of
uncertainty interact to affect social learning.  We reviewed and distilled the several empirical and
theoretical models in the literature along four uncertainty dimensions: temporal
environmental uncertainty, selection set size, ambiguous payoffs, and effective lifespan.
The model reproduced classic predictions that social learning can yield higher payoffs
than individual learning when uncertainty is sufficiently limited. Our model also
demonstrated a new insight into social learning as a scaffold for individual learning,
enabled by endowing agents with an evolutionarily- and biologically-plausible individual
learning mechanism, softmax search. Specifically, we found that agents could recover from
receiving outdated information, so social learning could evolve even though social
information was outdated more often than not. We found increased payoff ambiguity and
lifespan led to greater drift, which is essentially evolutionary uncertainty in a
finite-size population. Detailed theoretical models and
predictions of cultural evolution under uncertainty are critical for understanding how
humans adapt to an uncertain and rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, especially in the most vulnerable
communities~\cite{McNamara2020}, and to capitalize on new behavioral opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

@@

\section{Appendix}

\mt{TODO: Write supporting material explaining supplemental results}

@@

\begin{table}[h] \caption{Max. iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} 
    \toprule $B$ & \# not at fixation (out of 132k) & \% not fixated \\ 
    \midrule  2  & 192  & 0.1 \% \\ 
              4  & 274  & 0.2 \% \\ 
              10 & 2935 & 2.0 \% \\ 
    \bottomrule \end{tabular} 
\end{table}

\subsection{Full net payoff comparison for all nine uncertainty conditions}

>>>>

By disambiguating and organizing various operationalizations of uncertainty
into a single computational model, we developed a more nuanced theoretical
understanding of how different forms of uncertainty interact to affect the
evolution of social learning.  We reviewed and distilled several empirical
and theoretical models in the literature along four uncertainty dimensions:
temporal environmental uncertainty, selection set size, ambiguous payoffs, and
effective lifespan.  The model reproduced classic predictions that social
learning can yield higher payoffs than individual learning when uncertainty is
sufficiently limited. Our model also demonstrated a new insight into social
learning as a scaffold for individual learning, enabled by endowing agents
with an evolutionarily- and biologically-plausible individual learning
mechanism, softmax search. Specifically, we found that agents could recover
from receiving outdated information, so social learning could evolve even
though social information was outdated more often than not. We found increased
payoff ambiguity and lifespan led to greater drift, which is essentially
evolutionary uncertainty in a finite-size population. Detailed theoretical
models and predictions of cultural evolution under uncertainty are critical
for understanding how humans adapt to an uncertain and rapidly changing world,
both to mitigate existential threats~\cite{Moya2020,Jones2021}, especially in
the most vulnerable communities~\cite{McNamara2020}, and to capitalize on new
behavioral opportunities such as transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

@@

\section{Additional main figure: Full net payoff comparison for all nine uncertainty conditions}

In the main text we showed a $2\times2$ array of plots comparing the observed
mean net payoffs obtained by agents in our simulations, and in two auxiliary
sets of simulations that set all agents to either have the social learning
trait or not, which remained the case for the entire simulation since there
are no mutations in this model (Figure~\ref{fig:payoffs}). Here we show the full $3\times3$ array of
all $(\pilow, B)$ combinations shown for the main results. We see that
populations of all social learners especially underperform initially mixed
populations that evolve to be social learners 
(Figure~\ref{fig:fullMeanPrevNetPayoffs}, bottom row). This highlights the
importance of diversity for behavioral optimization, even if it is transient.

\section{Supplemental supporting information}

In this Supplement we support our model conclusions with auxiliary data
showing, first, sufficient model convergence and, second, that our model
outcomes are stable over various settings of the three free auxiliary model
parameters: the number of agents ($N$), the number of
prospective teachers pooled by social learner children for social learning
($N_T$), and the softmax behavior selection greediness parameter ($\beta$).
The main source of minor deviations from the main results was greater drift
due to (1) finite population size effects, with greater drift for smaller $N$;
(2) lower quality of socially-learned information agents can access with fewer
prospective teachers; and (3) lower quality social information when $\beta$
was small since teacher agents would have explored non-optimal behaviors more
frequently, so their observed payoffs would have less probability weight in
the optimal payoffs on average. We give details below of model convergence, and
outcomes as we varied the three auxiliary parameters $N$, $N_T$, and $\beta$.

@@

Model populations reached fixation between few to 140 generations on
average (Figure~\ref{fig:steps}).
Note 140 generations is equivalent to 2800 time steps. We set the maximum 
number of iterations globally to $20~000$, which is 1000 generations when
$L=20$. Although we can be relatively sure that this maximum number of 
generations is rarely reached based on our main text analysis of $\meanG$,
we additionally tabulated the number of simulations that did not converge to
fixation
within this step limit. We found that 0.1\% of simulations did not converge
when $B=2$, 0.2\% did not converge when $B=4$, and 2.0\% did not converge
when $B=10$ (Table~\ref{tab:convergence}). When a trial did not converge we
still included that trial's $\meansl$ value in the average across trials,
but for these $0 < \meansl < 1$.

\begin{table}[h] \caption{Maximum number of iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} 
    \toprule $B$ & \# not at fixation (out of 132k) & \% not fixated \\ 
    \midrule  2  & 192  & 0.1 \% \\ 
              4  & 274  & 0.2 \% \\ 
              10 & 2935 & 2.0 \% \\ 
    \bottomrule \end{tabular} 
\end{table}
